Whole-chromosome pairwise alignment using a hierarchical wavefront algorithm

### 2021-06-12 manuscript/main.tex
little fixes; notes

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present \wfmash\, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely available at
        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping
    \citep{22388286, BWA_MEM, 23103880} to variant detection~\citep{21478889}, as well as de~novo
    assembly \citep{19251739} and pangenome building~\citep{33177663, 33066802}. Pairwise sequence alignment
    can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. This opportunity, in combination with the massive amount
    of data available, made possible thanks to the advancement in sequencing technology, presses for the
    development of tools able to compute pairwise alignment of very long sequences.

    To face the upcoming challenges in pangenomes building, we have developed \wfmash\, a new gap-affine
    pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
    and gap-affine penalties for insertion and deletions. \wfmash\ scales efficiently to large eukaryotic
    chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise
    alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the integration of
    MashMap~\citep{30423094}, a fast approximate aligner for computing local alignment boundaries between
    long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine algorithm
    that takes advantage of homologous regions between the sequences to accelerate the alignment process.

@@

    Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, fixing bugs
    and incorporating the robust winnowing \citep{Schleimer S. et al.  (2003)} in the minimizer downsampling.
    Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
    runtime and memory-usage without affecting accuracy~\citep{32657365}.

@@

    To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment
    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
    sequences, it aligns them to each other in small pieces W-bps long by applying a global WFA alignment
    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
    two fragments.

@@

    Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments
    on the optimal path in the high order DP-matrix. During the process, a further step is performed for
    refining the alignment. Indeed, inaccurate mapping estimation can lead to under-alignment at the beginning
    and the end of the aligned syntenic region. Therefore, edlib is applied by computing semi-global alignments
    to resolve the mapping boundaries.
    The hierarchical implementation requires only the memory to align a syntenic region at \textit{W}-bps resolution,
    limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
    This approach is more flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
    no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
    %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
    %TODO 'and can benefit from parallel exploration of the wavefront.' Can we?

@@

    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 {29750242} and Winnomap 2.03 {XXX}, in
    terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
    data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
    The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K
    CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.

    \subsection{Simulated data}
    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and
    the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of divergence and length
    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa}. In each
    run, all sequences have the same length. Single nucleotide variants (SNVs), small indels, and structural
    variants (SVs) were introduced using the Mutation-Simulator script~\citep{32780803}.
    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.

    To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences to build
    a pangenome graph in each run, and called the variants on the graphs. The graphs were built using~\citep{seqwish}.
    seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
    sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
    evaluating the false-negative and false-positive rates (FNR, FPR) using vcfeval~\citep{vcfeval}.

@@

@@

    We implemented a novel gap-affine pairwise aligner, \wfmash\, to accelerate the computation of the mutual
    alignment of collections of genomes, a step required for building pangenome models. We have demonstrated that
    it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
    successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise
    alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
    increasing yields of sequencing technologies in the coming years.

>>>>

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present \wfmash\/, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely available at
        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping
    \citep{22388286, BWA_MEM, 23103880} to variant detection~\citep{21478889}, as well as de~novo
    assembly \citep{19251739} and pangenome building~\citep{33177663, 33066802}. Pairwise sequence alignment
    can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. FIXME In combination with the massive amount
    of data available made possible thanks to the advancement in sequencing technology, presses for the
    development of tools able to compute pairwise alignment of very long sequences.

    To face the upcoming challenges in pangenomes building, we have developed \wfmash\/, a new gap-affine
    pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
    and gap-affine penalties for insertion and deletions. \wfmash\ scales efficiently to large eukaryotic
    chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise
    alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the integration of
    MashMap~\citep{30423094}, a fast approximate aligner for computing local alignment boundaries between
    long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine algorithm
    that takes advantage of homologous regions between the sequences to accelerate the alignment process.

@@

    Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, incorporating
    the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
    Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
    runtime and memory-usage without affecting accuracy~\citep{32657365}.

@@

    To avoid such limitation, wfmash applies a hierarchical WFA, exploiting the WFA to guide the alignment
    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
    two fragments.

@@

    Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments
    on the optimal path in the high order DP-matrix. During the process, a further step is performed for
    refining the alignment. Indeed, inaccurate mapping estimation can lead to under-alignment at the beginning
    and the end of the aligned syntenic region. Therefore, edlib is applied by computing semi-global alignments
    to resolve the mapping boundaries.

    The hierarchical implementation requires only the memory to align a syntenic region at \textit{W}-bps resolution,
    limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
    This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
    no heuristic seeding step, and FIXME can benefit from parallel exploration of the wavefront.
    %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
    %TODO 'and can benefit from parallel exploration of the wavefront.' Can we?

@@

    We evaluated the performance of \wfmash\ 0.xx against Minimap2-v2.20-r1061 \citep{29750242} and Winnomap 2.03 \citep{XXX}, in
    terms of runtime and memory usage. The comparison was performed on both simulated and real datasets. Simulated
    data allowed evaluating the alignment quality by calculating precision and recall in variant identification.
    The evaluations were performed on a workstation running Ubuntu 20.04 equipped with 64 GB of RAM and an Intel i9-9900K
    CPU with 8 cores (16 threads). All commands are provided in the Supplementary Material.

    \subsection{Simulated data}
    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and
    the telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of divergence and length
    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool~\citep{splitfa},
    In each run, all sequences have the same length, with 50\% of them in reverse complement.
    Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced using the
    Mutation-Simulator script~\citep{32780803}.
    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.

    To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences to build
    a pangenome graph in each run, identifying the variants embedded in the graphs. The graphs were built using~\citep{seqwish}.
    seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
    sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
    evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval}.

@@

    % TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000)

    % TODO Centromere (chr8:42,881,543-47,029,467)

    % TODO Simulation with full chromosomes

@@

    We implemented a novel gap-affine pairwise aligner, \wfmash\/, to accelerate the computation of the mutual
    alignment of collections of genomes, a step required for building pangenome models. We have demonstrated that
    it efficiently performs with contigs representing full human chromosomes of 88 phased haplotypes from the
    Human Pangenome Reference Consortium year 1 assembly. Indeed, thanks to its efficiency, wfmash is already
    successfully applied in our pangenome graphs building pipeline (\citep{pggb}). No less important, pairwise
    alignment is a central step of many bioinformatics applications, making our aligner a scalable solution to face the
    increasing yields of sequencing technologies in the coming years.
